---
title: "Modelled regional income timeseries"
author: "Alan Bentley"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
---
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = F, message=F, warning=F, cache=F)

library(dplyr)
library(ggplot2)
library(scales)
library(tidyr)
library(stringr)
library(zoo)
library(knitr)
library(kableExtra)
library(lubridate)

#Some colours
palc <- c("grey20", "#d2ac2f", "grey60", "#a4343e", "#35345d", "#76a93f", "#6f2e38", "#0d94a3", "#dd6829", "#1a6e5b")
palc2 <- c("grey20", "#d2ac2f","#00a0d5","#a4343e", "grey60","#35345d", "#76a93f", "#6f2e38", "#0d94a3", "#dd6829", "#1a6e5b")

# Custom ggplot (chart) themes
theme_ab = function(base_size=14) {
  theme_minimal() +
    theme(text = element_text(size = base_size, color="grey30"),
          plot.title = element_text(size=14),
          plot.subtitle = element_text(size=11),
          plot.caption = element_text(size=10),
          legend.text = element_text(size=10),
          axis.text.y = element_text(size=9),
          axis.text.x = element_text(size=9),
          axis.title = element_text(size=10),
          strip.text = element_text (color="grey30", size=9, margin = margin(0,.5,0,.5, "cm")),
          panel.spacing = unit(1, "lines"))
}

# Some helpful data functions

date_to_qtr <- function(dt){
  # Take a date and return the last day of the quarter that the date is in

  qtr_first_day <- yq(paste(year(dt), quarter(dt)))

  qtr_last_day <- ceiling_date(qtr_first_day, unit = "quarter") - days(1)

  qtr_last_day

}

date_to_month <- function(dt){
  # Take a date and return the last day of that month as a date

  ceiling_date(dt, unit = "months") - days(1)

}

date_to_week <- function(dt){
  # Take a date and return the last day of that week as a date (week ending Sunday)
  
  ceiling_date(dt, unit = "week", week_start = getOption("lubridate.week.start", 1)) - days(1)
  
}

```


\newpage

# Introduction

## Data needs

*Regional income* series -- split by regional council and territorial authority -- are needed to understand households' and individuals' economic resources. For example, we can use these series alongside rentals for housing and house prices to provide indicators of housing affordability.

Regional timeseries of household income are not readily available as official statistics. This document describes our modelling approach, which combines several data sources, to estimate regional household income.

## Modelling approach

Our modelling strategy is to use timely, but less comprehensive, indicators, calibrated to reliable, but infrequent, benchmarks available with a notable lag. This is a three stage approach:

1. Create **income benchmarks**: The Household Economic Survey (HES) is used to estimate *calibration and coverage factors* to apply to Census income estimates
2. **Interpolate** *between* income benchmarks: Linked employer-employee (LEED) tax data is used to estimate regional income timeseries between benchmarks (created in step 1)
3. **Extrapolate** *beyond* income benchmarks: LEED and other tax data is to used to extend the regional income timeseries to the latest period 

The approach is flexible so can be used to estimate several aggregate statistics:

* Gross (before tax) and disposable (after tax) income
* For various income sharing units: personal, household, equivalised household (accounts for different household composition)
* By *tenure* (renters and all households)
* For different *periodicities* (annual, quarterly)

\newpage

## Data used

Four income data sources are used: two surveys and two variants of administrative tax data (see Table 1). Individually, each source has critical weaknesses. Collectively, we can borrow strength from each source to model timely, reliable, estimates.

### Table 1: Summary of data sources {-}

| **Source**                | **Availability**         |**Quality characteristics**           |
| ------------------------- |:---------------:| :-----------------------:| 
| Household Economic Survey (**HES**)  | Yearly, reporting lag of about 18 months | Comprehensive coverage of income sources (gross and disposable); limited regional coverage as sample survey | 
| **Census** of population and dwellings  | 5-yearly, reporting lag of about 18 months | Comprehensive regional coverage; only gross (before tax) income - overall quality thought to be lower than HES | 
| Tax data: Linked employer-employee data (**LEED**)  | Quarterly, 4-5 quarter lag | Good regional coverage; *median gross earnings per job, based on employment location* (rather than home address) | 
| Tax data: **Business employment data**  | Quarterly, 2 month lag | Good regional coverage; *mean* gross earnings per job, based on *employer location* (rather than home address) | 

\newpage

# Income benchmarks

The **first stage** uses survey data; namely, the Census of population and dwellings, and Household Economic Survey (HES). The Census has comprehensive regional coverage (for example, series are available for territorial authorities and Auckland local boards), but only gross (before tax) income is reported. Census income is collected as discrete, banded, amounts, so is less precise than income reported as dollars. The quality of HES income series is considered superior as detailed collection, by income source (in dollars), aids coverage of income sources and allows for the calculation of disposable (after tax) income.

### Figure 1 {-}

```{r}
#1. Create coverage and calibration factors

#1.1 Read and tidy customised data 

#1.1.1 HES (available at https://catalogue.data.govt.nz/dataset/hes-gross-and-disposable-income-timeseries-2006-07-2018-19)

# Household income

HES_hh <- read.csv("https://catalogue.data.govt.nz/dataset/b94e07c7-b877-4a49-8eb7-6171e13f81aa/resource/048bf4ac-45db-4bba-b015-9b9226bc690d/download/hes-householdincome.csv") %>% 
  mutate(unit="Household") %>% 
  select(-c("MeanOECModIncome","MedianOECModIncome"))

# Personal income

HES_per <- read.csv("https://catalogue.data.govt.nz/dataset/b94e07c7-b877-4a49-8eb7-6171e13f81aa/resource/fed6d2ed-b07a-40b3-91bd-3866e0d686df/download/hes-personincome.csv") %>% 
  mutate(unit="Personal")



#Stats NZ data 2007 - 2021 (Revised data)

download.file("https://www.stats.govt.nz/assets/Uploads/Household-income-and-housing-cost-statistics/Household-income-and-housing-cost-statistics-Year-ended-June-2021/Download-data/household-income-and-housing-cost-statistics-year-ended-june-2021-CSV.zip",
              destfile = "housing_costs_2021.zip")

HES_inc_NEW <- read.csv(unz("housing_costs_2021.zip","inc_t3_datafile_2021.csv")) %>%
  filter( MsCode %in% c("M002"), #median
          IncCode %in% c("I001A","I002A")) %>% #gross or disposable
  select(Estimate,IncCode, Year) %>%
  mutate(date=as.Date(paste0(Year,"-03-31")),
         inc_type=ifelse(IncCode=="I001A","Gross","Disposable"),
         value=Estimate,
         tenure="All",
         unit="Household") %>%
  select(date,inc_type,value, tenure, unit) %>% 
  group_by(inc_type,tenure,unit) %>% 
  mutate(x=1:n(),
         value_smooth=predict(loess(value ~ x, span=0.5))) %>% 
  select(-x) %>% 
  gather(trend_type,value,-c(unit,tenure,inc_type,date)) %>% 
  mutate(trend_type=recode(trend_type,
                                  'value'="actual",
                                  'value_smooth'="smoothed"))

HES_inc <- HES_hh %>% 
  rbind(HES_per) %>% 
  gather(inc_type,value,-c(Year,RegCouncil,HouseholdTenure,unit)) %>% 
  filter(RegCouncil == "Total" &
           Year != "2019") %>% 
  group_by(inc_type,HouseholdTenure,unit) %>% 
  mutate(date=as.Date(paste0(Year,"-03-31")),
    x=1:n(),
         value_smooth=predict(loess(value ~ x, span=0.5)),
    tenure=recode(HouseholdTenure,
                  "Dwelling not owned, rent payments made"="Renters",
                  "Total"="All")) %>% #smooth series
  filter(inc_type %in% c("MedianTotalRegular","MedianDispIncome")) %>% 
  ungroup() %>% 
  select(unit,tenure,inc_type,value,date,value_smooth) %>% 
  gather(trend_type,value,-c(unit,tenure,inc_type,date)) %>% 
  mutate(trend_type=recode(trend_type,
                                  'value'="actual",
                                  'value_smooth'="smoothed"),
         inc_type=recode(inc_type,
                         "MedianTotalRegular"="Gross",
                         "MedianDispIncome"="Disposable")) %>% 
  filter(!(unit =="Household" & tenure =="All")) %>% #Over write old estimate with revised timeseries
  rbind(HES_inc_NEW)


#1.1.2 Census (available at https://catalogue.data.govt.nz/dataset/census-household-and-personal-income-series)

Census_hh <- read.csv("https://catalogue.data.govt.nz/dataset/c2467785-a23b-4517-8be9-5ffc1700635c/resource/f9668f69-f6eb-42e5-97b5-ab04eafeab53/download/t1-hhincome-oecdmod-2001-2018.csv") %>% 
  filter(RegionalCouncil == "Total") %>% 
  mutate(date=as.Date(paste0(Year,"-03-31"))) %>% 
  rename(tenure="HouseholdTenure") %>% 
  gather(unit,value,-c(Year,date,tenure)) %>% 
  filter(tenure %in% c("Total","21 Rented"),
         unit %in% c("MedianHouseholdIncome","MedianPersonalIncome")) %>% 
  mutate(value=as.numeric(value),
         tenure = recode(tenure, "21 Rented"="Renters",
                         "Total"="All"),
         unit=recode(unit,
                     "MedianHouseholdIncome"="Household",
                     "MedianPersonalIncome"="Personal"),
         inc_type = "Gross") 

# 1.2 Plot the data

p1 <- ggplot(HES_inc, aes(date,value))+
  geom_line(aes(size=trend_type, colour=inc_type))+
  geom_point(data=Census_hh, aes(shape="Census"), size=2, colour=palc[2])+
  facet_grid(unit~tenure)+
  theme_ab()+
  scale_size_manual(values=c(.3,.8), labels=c("actual"="HES, actual","smoothed"="HES, smoothed"), name="")+
  scale_colour_manual(values=palc,name="")+
  scale_linetype_manual(values=c("dotted","solid"), name="")+
  scale_shape_manual(values=19, name="")+
  scale_y_continuous(labels=dollar)+
  labs(title="Comparison of median income estimates",
       subtitle="HES and Census, by tenure and income sharing unit",
       y="", x="",
       caption="Thin lines are actual HES estimates,
       Thick lines are smoothed HES estimates (LOESS trends) \n
       Source: Stats NZ and author's calculations")+
  guides(size = guide_legend(order=1),
         shape = guide_legend(override.aes = list(colour = palc[2]), order=2),
         colour = guide_legend(override.aes = list(size = 5)), order=3)

p1


```


Shown in Figure 1, gross income is usually a higher estimate using the HES compared with Census (using revised HES[^1], published in February 2020, household income for all tenures is slightly lower in 2013 using HES, but was higher in 2018, compared with comparable Census estimates). We smoothed the HES timeseries using a LOESS (locally estimated scatterplot smoothing) regression to reduce HES sampling errors. The HES series overlap with the the 2013 and 2018 Censuses (the income reference period is assumed comparable: Censuses are conducted in March; HES is conducted continuously, reported for each year to June). Calculating the average ratio of HES to Census gross income estimates produces gross income *calibration factors* (see Table 2 and Figure 2).  These have been calculated for household and personal income, by tenure type (renters and all households).

[^1]: Revisions to the HES income estimator are discussed in the appendix. Estimates of different vintages are shown in appendix Figure A1.

Replacing HES gross income with disposable income we can calculate analogous *calibration and coverage factors* for disposable income.

\newpage


### Table 2: Census calibration and coverage factors {-}


```{r}
#2.1. calculate census calibration and coverage factor

Census_hh2 <- Census_hh %>% 
  filter(date %in% c(as.Date("2013-03-31"),as.Date("2018-03-31"))) %>% 
  select(-c(Year,inc_type)) %>% 
  rename(cvalue=value)

c_factor <- HES_inc %>% 
  filter(trend_type == "smoothed",
         date %in% c(as.Date("2013-03-31"),as.Date("2018-03-31"))) %>% 
  select(-trend_type) %>% 
  right_join(Census_hh2, by=c("unit","tenure","date")) %>% 
  mutate(cfactor=value/cvalue) %>% 
  group_by(unit,tenure,inc_type) %>% 
  summarise(cfactor=(mean(cfactor))) %>% 
  ungroup()

c_factor_tab <- c_factor %>%
  mutate(cfactor=round(cfactor,2)) %>% 
  pivot_wider(names_from=inc_type,values_from=cfactor)

kable(c_factor_tab,
      booktabs = T, col.names = c("Income unit","Tenure","Disposable","Gross")) %>% 
  add_header_above(c(" "=2, "Income type" = 2)) %>% 
  kable_styling(position = "center")

```

### Figure 2 {-}

```{r}

#2.2 plot calibration and coverage factors

HES_inc2 <- filter(HES_inc, unit=="Household", tenure=="All", trend_type=="smoothed")

Census_hh2 <- filter(Census_hh, unit=="Household", tenure=="All")

p2 <- ggplot(HES_inc2, aes(date,value, colour=inc_type))+
  geom_line()+
  geom_point(data=filter(HES_inc2, date %in% c(as.Date("2013-03-31"),as.Date("2018-03-31"))), aes(shape="HES"), size=3)+
  geom_curve(aes(x=as.Date("2013-03-31"),xend=as.Date("2013-03-31"),y=63800,yend=	
63399), curvature = 0.5, size=.5, colour=palc[2], arrow = arrow(length = unit(0.03, "npc")))+
  geom_curve(aes(x=as.Date("2018-03-31"),xend=as.Date("2018-03-31"),y=75700,yend=78843.03), curvature = 0.5, size=.5, colour=palc[2], arrow = arrow(length = unit(0.03, "npc")))+
  geom_curve(aes(x=as.Date("2013-03-31"),xend=as.Date("2013-03-31"),y=63800,yend=55120.18), curvature = 0.5, size=.5, colour=palc[4], arrow = arrow(length = unit(0.03, "npc")))+
  geom_curve(aes(x=as.Date("2018-03-31"),xend=as.Date("2018-03-31"),y=75700,yend=66752.36), curvature = 0.5, size=.5, colour=palc[4], arrow = arrow(length = unit(0.03, "npc")))+
  geom_point(data=Census_hh2, aes(shape="Census"), size=3)+
  annotate("text", x = as.Date("2012-03-31"), y = 77000, label = "gross income calibration factor", colour=palc[2])+
  annotate("text", x = as.Date("2015-03-31"), y = 47000, label = "disposable \nincome \ncalibration and \ncoverage factor", colour=palc[4])+
  # facet_grid(unit~tenure, scales="free")+
  theme_ab()+
  scale_size_manual(values=c(.1,1), name="")+
  scale_colour_manual(values=c("grey70","grey40"),name="")+
  scale_linetype_manual(values=c("dotted","solid"), name="")+
  scale_shape_manual(values=c(19,1), name="")+
  scale_y_continuous(labels=dollar)+
  labs(title="Calibration and coverage factors: calculation",
       subtitle="All tenures, household income",
       y="", x="",
       caption="Smoothed HES estimates (LOESS trends) \n
       Source: Stats NZ and author's calculations")+
  theme(panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())

p2

```


\newpage

### Figure 3 {-}

The resulting New Zealand level benchmark estimates, after applying the calibration and coverage factors, to Census data, is shown in Figure 3. The same factors were applied for each regional area of New Zealand.

```{r}
#3. Create benchmarks

#3.1 Census for all areas 

Census_hh <- read.csv("https://catalogue.data.govt.nz/dataset/c2467785-a23b-4517-8be9-5ffc1700635c/resource/f9668f69-f6eb-42e5-97b5-ab04eafeab53/download/t1-hhincome-oecdmod-2001-2018.csv", stringsAsFactors = F) %>%  
  filter(RegionalCouncil == "Total") %>% 
  mutate(date=as.Date(paste0(Year,"-03-31"))) %>% 
  rename(tenure="HouseholdTenure") %>% 
  gather(unit,value,-c(Year,date,tenure)) %>% 
  filter(tenure %in% c("Total","21 Rented"),
         unit %in% c("MedianHouseholdIncome","MedianPersonalIncome")) %>% 
  mutate(value=as.numeric(value),
         tenure = recode(tenure, "21 Rented"="Renters"),
         unit=recode(unit,
                     "MedianHouseholdIncome"="Household",
                     "MedianPersonalIncome"="Personal"),
         inc_type = "Gross")

#3.2 Create combo file from separate area level files

cc1 <- read.csv("https://catalogue.data.govt.nz/dataset/c2467785-a23b-4517-8be9-5ffc1700635c/resource/f9668f69-f6eb-42e5-97b5-ab04eafeab53/download/t1-hhincome-oecdmod-2001-2018.csv", stringsAsFactors = F) %>% 
  rename(area=RegionalCouncil) %>% 
  mutate(area=ifelse(area=="Total","New Zealand",area),
         area=str_trim(gsub('[0-9]+', '', area)),
    level = ifelse(area=="New Zealand","New Zealand","Regional Council"),
    area = ifelse(area=="New Zealand","New Zealand",paste0(area," Region")))

cc2 <- read.csv("https://catalogue.data.govt.nz/dataset/c2467785-a23b-4517-8be9-5ffc1700635c/resource/dc7e26dc-eba7-43f9-b2c2-854494c38543/download/t2-ta-hhincome-oecdmod-2001-2018.csv", stringsAsFactors = F) %>%  
  rename(area=TerritorialAuthority) %>% 
  mutate(level="Territorial Local Authority")

cc3 <- read.csv("https://catalogue.data.govt.nz/dataset/c2467785-a23b-4517-8be9-5ffc1700635c/resource/2ee674bd-eb7c-4e40-8ed0-bedf7c44b059/download/t3-alb-hhincome-oecdmod-2001-2018.csv", stringsAsFactors = F) %>% 
  rename(area=LocalBoardArea) %>% 
  mutate(level="Auckland Local Board") %>% 
  filter(area !="Total")

cc <- rbind(cc1, cc2, cc3) %>% 
  filter(area != "99 Area outside region") %>% 
  mutate(date=as.Date(paste0(Year,"-03-31"))) %>% 
  rename(tenure="HouseholdTenure") %>%
  select(-Year) %>% 
  gather(stat,value,-c(date,area,level,tenure)) %>% 
  filter(tenure %in% c("Total","21 Rented"),
         grepl("Median",stat)) %>% 
  mutate(value=as.numeric(value),
         tenure = recode(tenure, "21 Rented"="Renters",
                         "Total"="All"),
         unit=ifelse(stat == "MedianPersonalIncome", "Personal", "Household"),
         unit2=recode(stat,
                     "MedianPersonalIncome"="Personal",
                     "MedianHouseholdIncome"="Household",
                     "MedianOECDModifiedIncome"="Equivalised Household")) %>% 
  right_join(c_factor_tab, by=c("unit","tenure")) %>% 
  mutate(value_gross=value*Gross,
         value_disposable=value*Disposable) %>% 
  select(-c("stat","unit","value","Gross","Disposable")) %>% 
  rename(unit=unit2) %>% 
  gather(inc_type,value,-c(area,tenure,level,date,unit)) %>% 
  mutate(inc_type=recode(inc_type,
                         "value_gross"="Gross",
                         "value_disposable"="Disposable"),
         source="Census") %>% 
  filter(!(area %in% c("Area outside region Region","Area Outside Territorial Authority",
                       "Chatham Islands Territory","Auckland","Total")) &
           level != "Auckland Local Board")
  

NZ_benchmarks <- filter(cc, level=="New Zealand")

Reg_benchmarks <- filter(cc, level=="Regional Council")

TA_benchmarks <- filter(cc, level=="Territorial Local Authority")

#3.3 Plot the NZ benchmarks

p3 <- ggplot(NZ_benchmarks, aes(date,value, colour=inc_type))+
  geom_point(size=2)+
  facet_grid(tenure~unit)+
  theme_ab()+
  scale_colour_manual(values=palc,name="")+
  scale_y_continuous(labels=dollar)+
  labs(title="Benchmark values",
       subtitle="New Zealand, Income sharing unit by tenure",
       y="", x="",
       caption="Source: Stats NZ and author's calculations")+
  theme(legend.position="top")

p3

```

\newpage

# Interpolation and extrapolation

## Method explained

The **second stage** uses Linked employer-employee (LEED) tax data to **interpolate** *between* income benchmarks.

1. LEED median gross earnings per job are converted from *quarterly* to *annual* amounts, by aggregating over a rolling 4-quarter period.
2. *Benchmark factors* are calculated as the ratio of annualised LEED median earnings to income benchmarks (estimated from HES and Census in stage 1)
3. A timeseries of benchmark factors (between Census years) is created by linear interpolation
4. Applying these factors to the LEED earnings timeseries results in calibrated income timeseries

The **third stage** uses LEED and other tax data to extend the timeseries to the latest period.

1. LEED quarterly series are available with a 4-5 quarter lag. The latest available benchmark factor (from the time of the most recent Census) is applied to the LEED earnings series available beyond the benchmark.
2. This series is further extended to the latest period using annual changes (on a quarterly frequency) in another source of tax data, *Business employment data*, mean gross earnings are now used as median series are currently unavailable.

## Final income timeseries

The resulting timeseries, for total New Zealand, are shown in Figure 4. The same methodology is applied for each regional area of New Zealand.

### Figure 4 {-}

```{r, fig.height=6}
#4. Interpolate and extrapolate benchmarks with tax data (ANNUAL income)

#4.1 Read LEED (Linked Employer-Employee Data) from NZ.Stat

# https://nzdotstat.stats.govt.nz/wbos/Index.aspx?DataSetCode=TABLECODE7037

leed_ta <- read.csv("../data/LEED_table37.csv", na.strings=c("","NA"), stringsAsFactors = F)[,c(1:2,7)] %>% 
  fill(Quarter) %>%
  rename(area=Territorial.authority,
         med_inc=Median.earnings.of.full.quarters.jobs) %>% 
  filter(med_inc!="..") %>% 
  mutate(area=str_trim(area),
         area=gsub("district","District",area),
         area=gsub("city","City",area),
         area=gsub(" \\(modified\\)","",area),
         area=gsub("Total territorial authority","New Zealand",area),
         area=gsub("Wanganui District","Whanganui District",area),
         med_inc=as.numeric(med_inc),
         date=as.Date(paste0(Quarter,"-01"),format="%b-%y-%d"),
         date=date_to_qtr(date),
         source="LEED",
         level=ifelse(area=="New Zealand","New Zealand","Territorial Local Authority")) %>% 
  select(-Quarter) %>% 
  group_by(area) %>% 
  arrange(date) %>% 
  mutate(med_inc=med_inc+lag(med_inc,1)+lag(med_inc,2)+lag(med_inc,3)) %>% #annual income
  ungroup()

# https://nzdotstat.stats.govt.nz/wbos/Index.aspx?DataSetCode=TABLECODE7005

leed_reg <- read.csv("../data/LEED_table5.csv", na.strings=c("","NA"), stringsAsFactors = F)[,c(1:2,7)] %>% 
  fill(Quarter) %>%
  rename(area=Region,
         med_inc=Median.earnings.of.full.quarters.jobs) %>% 
  filter(med_inc!="..") %>% 
  mutate(area=paste0(str_trim(area), " Region"),
         area=gsub("Manawatu-Wanganui Region","Manawatu-Whanganui Region",area),
         med_inc=as.numeric(med_inc),
         date=as.Date(paste0(Quarter,"-01"),format="%b-%y-%d"),
         date=date_to_qtr(date),
         source="LEED",
         level="Regional Council") %>% 
  select(-Quarter) %>% 
  group_by(area) %>% 
  arrange(date) %>% 
  mutate(med_inc=med_inc+lag(med_inc,1)+lag(med_inc,2)+lag(med_inc,3)) %>% 
  ungroup() %>% 
  filter(area != "Total Region Region")

leed <- rbind(leed_ta, leed_reg) %>% 
  group_by(area) %>% #Remove incomplete series (Auckland old TAs)
  mutate(max_date=max(date),
         min_date=min(date)) %>%
  ungroup() %>% 
  filter(max_date==max(date),
         min_date==min(date)) %>% 
  select(-c(max_date,min_date)) %>% 
  rename(value=med_inc) %>%
  filter(date >= as.Date("2001-03-31")) #remove prior to 2001 Census

#4.2 More recent quarters from Business Employment Statistics

download.file("https://www.stats.govt.nz/assets/Uploads/Business-employment-data/Business-employment-data-March-2022-quarter/Download-data/business-employment-data-march-2022-quarter-csv.zip",destfile = "bus_emp_mar22.zip")
bus_emp <- read.csv(unz("bus_emp_mar22.zip", "machine-readable-business-employment-data-mar-2022-quarter.csv"), stringsAsFactors = F)

bus_emp2 <- bus_emp %>% 
  filter(Series_title_2 =="Total Industry"|
    Group == "Territorial authority by employment variable" & Series_title_2 !="Auckland"|
    Group %in% c("Region by employment variable"),
         Series_title_1 %in% c("Filled jobs", "Total earnings"),
         Series_title_2 != "Area Outside Territorial Authority",
        Series_title_3 == "Actual") %>% 
  mutate(Series_title_2 = ifelse(Group == "Region by employment variable", paste0(Series_title_2," Region"), Series_title_2)) %>% 
  select(Period,Data_value,Series_title_1,Series_title_2) %>% 
  mutate(date=as.Date(ifelse(as.character(substr(Period, 6, 7))!="1",paste0(Period,".01"),paste0(as.character(Period),"0.01")),"%Y.%m.%d"),
         date=date_to_month(date)) %>% 
  pivot_wider(names_from = Series_title_1, values_from = Data_value) %>% 
  select(-Period) %>% 
  setNames(c("area","date","jobs","earnings")) %>% 
   mutate(earnings=earnings*1000000,
    ave_earnings=(earnings/jobs), #average earnings, per quarter
    area=recode(area, 'Total Industry'="New Zealand")) %>% 
  group_by(area) %>% 
  arrange(date) %>% 
  mutate(med_inc=ave_earnings+lag(ave_earnings,1)+lag(ave_earnings,2)+lag(ave_earnings,3), # annual
         source="Business employment data, mean earnings") %>% 
  select(date,area, med_inc,area,source) %>% 
  rename(value=med_inc) %>% 
  mutate(level=ifelse(grepl("Region",area),"Regional Council",
                      ifelse(area=="New Zealand","New Zealand","Territorial Local Authority")))

bus_emp3 <- bus_emp2 %>% 
  filter(date >= max(leed$date)) %>%
  arrange(date) %>% 
  mutate(index=value/value[1]) %>% 
  select(date,index,area, level) %>% 
  filter(date != max(leed$date))

#4.3 Interpolate HES/Census benchmarks with LEED and business employment data

polate <- function (unit_in, inc_type_in, tenure_in) {
  
  combo1 <- cc %>% 
  filter(unit==unit_in, inc_type==inc_type_in,
         tenure==tenure_in) %>% 
  select(-c(unit,inc_type,tenure)) %>% 
  rbind(leed) %>% 
  pivot_wider (names_from=source, values_from=value) %>% 
  group_by(area) %>%
  arrange(date) %>%
  mutate(ratio= Census/LEED, #calibration factor
         ratio=ifelse(is.na(ratio),na.approx(ratio),ratio), #interpolate between Censuses
         ratio=ifelse(date <= as.Date("2018-03-31"), ratio, ratio[date==as.Date("2018-03-31")]), #extrapolate beyond
         LEED2 = LEED*ratio)

combo2 <- rbind(combo1,bus_emp3) %>% 
  group_by(area) %>% 
  mutate(LEED3=ifelse(is.na(LEED2),LEED2[date==max(leed$date)]*index,LEED2),
         source=ifelse(date <= as.Date("2018-03-31"),"LEED interpolated",
                       ifelse(date > as.Date("2018-03-31") & date <= max(leed$date), 
                              "LEED extrapolated", "Tax extrapolated"))) %>% 
  rename(value=LEED3) %>% 
  ungroup() %>% 
  select(area,date,source,value) %>% 
  mutate(unit=unit_in, 
         inc_type=inc_type_in,
         tenure=tenure_in)

combo2

}

combo_polate <- c()

combo_polate <- rbind(combo_polate, polate(unit_in="Household", inc_type_in="Gross", tenure_in="All"))
combo_polate <- rbind(combo_polate, polate(unit_in="Household", inc_type_in="Disposable", tenure_in="All"))
combo_polate <- rbind(combo_polate, polate(unit_in="Personal", inc_type_in="Gross", tenure_in="All"))
combo_polate <- rbind(combo_polate, polate(unit_in="Personal", inc_type_in="Disposable", tenure_in="All"))
combo_polate <- rbind(combo_polate, polate(unit_in="Equivalised Household", inc_type_in="Gross", tenure_in="All"))
combo_polate <- rbind(combo_polate, polate(unit_in="Equivalised Household", inc_type_in="Disposable", tenure_in="All"))

combo_polate <- rbind(combo_polate, polate(unit_in="Household", inc_type_in="Gross", tenure_in="Renters"))
combo_polate <- rbind(combo_polate, polate(unit_in="Household", inc_type_in="Disposable", tenure_in="Renters"))
combo_polate <- rbind(combo_polate, polate(unit_in="Personal", inc_type_in="Gross", tenure_in="Renters"))
combo_polate <- rbind(combo_polate, polate(unit_in="Personal", inc_type_in="Disposable", tenure_in="Renters"))
combo_polate <- rbind(combo_polate, polate(unit_in="Equivalised Household", inc_type_in="Gross", tenure_in="Renters"))
combo_polate <- rbind(combo_polate, polate(unit_in="Equivalised Household", inc_type_in="Disposable", tenure_in="Renters"))

HUD_mod_inc_a <- combo_polate %>% 
  mutate(vintage=Sys.Date(),
         source=factor(source, levels=c("LEED interpolated","LEED extrapolated","Tax extrapolated")),
         inc_freq="annual",
         level=ifelse(grepl("Region",area),"Regional Council",
                      ifelse(area=="New Zealand","New Zealand","Territorial Local Authority")))


#4.4 Plot for NZ

HUD_mod_inc_NZ <- filter(HUD_mod_inc_a, area=="New Zealand") %>%
  group_by(tenure,unit) %>% 
  mutate(source_max=lead(date))

p4 <- ggplot(HUD_mod_inc_NZ, aes(date,value))+
  geom_rect(aes(xmin = date, xmax = source_max, ymin = -Inf, ymax = Inf, fill = source), alpha=.3)+
  geom_point(data=NZ_benchmarks, aes(shape="benchmarks",colour=inc_type), size=2)+
  geom_line(aes(colour=inc_type))+
  facet_grid(tenure~unit)+
  theme_ab()+
  scale_colour_manual(values=palc,name="")+
  scale_fill_manual(values=c("grey85","grey70","grey55"), name="")+
  scale_size_manual(values=c(.7,1,.2))+
  scale_linetype_manual(values="dotted")+
  scale_y_continuous(labels=dollar)+
  scale_shape_manual(values=c(19,3), name="")+
  labs(title="Modelled timeseries",
       subtitle="New Zealand, Income sharing unit by tenure",
       y="", x="",
       caption="Source: Stats NZ and author's calculations")+
  theme(legend.position="bottom")+
  guides(fill = guide_legend(ncol=1),
         colour = guide_legend(ncol=1), override.aes = list(alpha = 1))

p4

```

The method can also be used to create quarterly timeseries. LEED series now use the original quarterly values (not rolled up to annual amounts) and the Census-HES benchmarks are divided by four to estimate quarterly values. Shown in Figure 5, the series are more volatile, but changes in income growth may be apparent sooner using a quarterly income series.

### Figure 5 {-}


```{r, fig.height=6}
#5. Interpolate and extrapolate benchmarks with tax data (QUARTERLY income)

#5.1 Read LEED (Linked Employer-Employee Data) from NZ.Stat

leed_ta <- read.csv("../data/LEED_table37.csv", na.strings=c("","NA"), stringsAsFactors = F)[,c(1:2,7)] %>% 
  fill(Quarter) %>%
  rename(area=Territorial.authority,
         med_inc=Median.earnings.of.full.quarters.jobs) %>% 
  filter(med_inc!="..") %>% 
  mutate(area=str_trim(area),
         area=gsub("district","District",area),
         area=gsub("city","City",area),
         area=gsub(" \\(modified\\)","",area),
         area=gsub("Total territorial authority","New Zealand",area),
         area=gsub("Wanganui District","Whanganui District",area),
         med_inc=as.numeric(med_inc),
         date=as.Date(paste0(Quarter,"-01"),format="%b-%y-%d"),
         date=date_to_qtr(date),
         source="LEED",
         level=ifelse(area=="New Zealand","New Zealand","Territorial Local Authority")) %>% 
  select(-Quarter) %>% 
  group_by(area) %>% 
  arrange(date) %>% 
  mutate(med_inc=med_inc) %>% #quarterly income
  ungroup()

leed_reg <- read.csv("../data/LEED_table5.csv", na.strings=c("","NA"), stringsAsFactors = F)[,c(1:2,7)] %>%
  fill(Quarter) %>%
  rename(area=Region,
         med_inc=Median.earnings.of.full.quarters.jobs) %>% 
  filter(med_inc!="..") %>% 
  mutate(area=paste0(str_trim(area), " Region"),
         area=gsub("Manawatu-Wanganui Region","Manawatu-Whanganui Region",area),
         med_inc=as.numeric(med_inc),
         date=as.Date(paste0(Quarter,"-01"),format="%b-%y-%d"),
         date=date_to_qtr(date),
         source="LEED",
         level="Regional Council") %>% 
  select(-Quarter) %>% 
  group_by(area) %>% 
  arrange(date) %>% 
  mutate(med_inc=med_inc) %>%  #quarterly income
  ungroup() %>% 
  filter(area != "Total Region Region")

leed <- rbind(leed_ta, leed_reg) %>% 
  group_by(area) %>% #Remove incomplete series (Auckland old TAs)
  mutate(max_date=max(date),
         min_date=min(date)) %>%
  ungroup() %>% 
  filter(max_date==max(date),
         min_date==min(date)) %>% 
  select(-c(max_date,min_date)) %>% 
  rename(value=med_inc) %>%
  filter(date >= as.Date("2001-03-31")) #remove prior to 2001 Census

#5.2 More recent quarters from Business Employment Statistics

bus_emp2 <- bus_emp %>% 
  filter(Series_title_2 =="Total Industry"|
    Group == "Territorial authority by employment variable" & Series_title_2 !="Auckland"|
    Group %in% c("Region by employment variable"),
         Series_title_1 %in% c("Filled jobs", "Total earnings"),
         Series_title_2 != "Area Outside Territorial Authority",
    Series_title_3 == "Actual") %>% 
  mutate(Series_title_2 = ifelse(Group == "Region by employment variable", paste0(Series_title_2," Region"), Series_title_2)) %>% 
  select(Period,Data_value,Series_title_1,Series_title_2) %>% 
  mutate(date=as.Date(ifelse(as.character(substr(Period, 6, 7))!="1",paste0(Period,".01"),paste0(as.character(Period),"0.01")),"%Y.%m.%d"),
         date=date_to_month(date)) %>% 
  pivot_wider(names_from = Series_title_1, values_from = Data_value) %>% 
  select(-Period) %>% 
  setNames(c("area","date","jobs","earnings")) %>% 
   mutate(earnings=earnings*1000000,
    ave_earnings=(earnings/jobs), #average earnings, per quarter
    area=recode(area, 'Total Industry'="New Zealand")) %>% 
  group_by(area) %>% 
  arrange(date) %>% 
  mutate(med_inc=ave_earnings, # quarterly
         source="Business employment data, mean earnings") %>% 
  select(date,area, med_inc,area,source) %>% 
  rename(value=med_inc) %>% 
  mutate(level=ifelse(grepl("Region",area),"Regional Council",
                      ifelse(area=="New Zealand","New Zealand","Territorial Local Authority")))

bus_emp3 <- bus_emp2 %>% 
  filter(date >= max(leed$date)) %>%
  arrange(date) %>% 
  mutate(index=value/value[1]) %>% 
  select(date,index,area, level) %>% 
  filter(date != max(leed$date))

#5.3 Estimate census quarterly income

ccq <- cc %>% 
  mutate(value=value/4) # quarterly estimate

#5.4 Interpolate HES/Census benchmarks with LEED and business employment data

polate <- function (unit_in, inc_type_in, tenure_in) {
  
  combo1 <- ccq %>% 
  filter(unit==unit_in, inc_type==inc_type_in,
         tenure==tenure_in) %>% 
  select(-c(unit,inc_type,tenure)) %>% 
  rbind(leed) %>% 
  pivot_wider (names_from=source, values_from=value) %>% 
  group_by(area) %>%
  arrange(date) %>%
  mutate(ratio= Census/LEED, #calibration factor
         ratio=ifelse(is.na(ratio),na.approx(ratio),ratio), #interpolate between Censuses
         ratio=ifelse(date <= as.Date("2018-03-31"), ratio, ratio[date==as.Date("2018-03-31")]), #extrapolate beyond
         LEED2 = LEED*ratio)

combo2 <- rbind(combo1,bus_emp3) %>% 
  group_by(area) %>% 
  mutate(LEED3=ifelse(is.na(LEED2),LEED2[date==max(leed$date)]*index,LEED2),
         source=ifelse(date <= as.Date("2018-03-31"),"LEED interpolated",
                       ifelse(date > as.Date("2018-03-31") & date <= max(leed$date), 
                              "LEED extrapolated", "Tax extrapolated"))) %>% 
  rename(value=LEED3) %>% 
  ungroup() %>% 
  select(area,date,source,value) %>% 
  mutate(unit=unit_in, 
         inc_type=inc_type_in,
         tenure=tenure_in)

combo2

}

combo_polate <- c()

combo_polate <- rbind(combo_polate, polate(unit_in="Household", inc_type_in="Gross", tenure_in="All"))
combo_polate <- rbind(combo_polate, polate(unit_in="Household", inc_type_in="Disposable", tenure_in="All"))
combo_polate <- rbind(combo_polate, polate(unit_in="Personal", inc_type_in="Gross", tenure_in="All"))
combo_polate <- rbind(combo_polate, polate(unit_in="Personal", inc_type_in="Disposable", tenure_in="All"))
combo_polate <- rbind(combo_polate, polate(unit_in="Equivalised Household", inc_type_in="Gross", tenure_in="All"))
combo_polate <- rbind(combo_polate, polate(unit_in="Equivalised Household", inc_type_in="Disposable", tenure_in="All"))

combo_polate <- rbind(combo_polate, polate(unit_in="Household", inc_type_in="Gross", tenure_in="Renters"))
combo_polate <- rbind(combo_polate, polate(unit_in="Household", inc_type_in="Disposable", tenure_in="Renters"))
combo_polate <- rbind(combo_polate, polate(unit_in="Personal", inc_type_in="Gross", tenure_in="Renters"))
combo_polate <- rbind(combo_polate, polate(unit_in="Personal", inc_type_in="Disposable", tenure_in="Renters"))
combo_polate <- rbind(combo_polate, polate(unit_in="Equivalised Household", inc_type_in="Gross", tenure_in="Renters"))
combo_polate <- rbind(combo_polate, polate(unit_in="Equivalised Household", inc_type_in="Disposable", tenure_in="Renters"))

HUD_mod_inc_q <- combo_polate %>% 
  mutate(vintage=Sys.Date(),
         source=factor(source, levels=c("LEED interpolated","LEED extrapolated","Tax extrapolated")),
         inc_freq="quarterly",
         level=ifelse(grepl("Region",area),"Regional Council",
                      ifelse(area=="New Zealand","New Zealand","Territorial Local Authority")))

#5.5 Combine quarterly and annual data

HUD_mod_inc <- rbind(HUD_mod_inc_a, HUD_mod_inc_q) %>% 
  mutate(series=paste0("Median ",inc_freq," ", str_to_lower(unit), " ", str_to_lower(inc_type), " income for ", str_to_lower(tenure)))


#5.6 Plot for NZ

HUD_mod_inc_NZ <- filter(HUD_mod_inc, area=="New Zealand", inc_freq=="quarterly") %>%
  group_by(tenure,unit) %>% 
  mutate(source_max=lead(date))


p5 <- ggplot(HUD_mod_inc_NZ, aes(date,value))+
  geom_rect(aes(xmin = date, xmax = source_max, ymin = -Inf, ymax = Inf, fill = source), alpha=.3)+
  geom_point(data=NZ_benchmarks, aes(y=value/4, shape="benchmarks",colour=inc_type), size=2)+
  geom_line(aes(colour=inc_type), size=.2)+
  geom_smooth(se=F, aes(colour=inc_type), size=.8)+
  facet_grid(tenure~unit)+
  theme_ab()+
  scale_colour_manual(values=palc,name="")+
  scale_fill_manual(values=c("grey85","grey70","grey55"), name="")+
  scale_size_manual(values=c(.7,1,.2))+
  scale_y_continuous(labels=dollar)+
  scale_shape_manual(values=19, name="")+
  labs(title="Modelled timeseries: quarterly",
       subtitle="New Zealand, Income sharing unit by tenure",
       y="", x="",
       caption="Thin lines: original series
       Thick lines: LOESS trend
       \nSource: Stats NZ and author's calculations")+
  theme(legend.position="bottom")+
  guides(fill = guide_legend(ncol=1),
         colour = guide_legend(ncol=1), override.aes = list(alpha = 1))

p5

```


Median annual household disposable income by regional council is shown in Figure 6, and by territorial authority in Figure 7, as an examples of regional timeseries.


### Figure 6 {-}

```{r, fig.height=8}
#6. Plot of regional timeseries

HUD_mod_inc_reg <- filter(HUD_mod_inc, level=="Regional Council", inc_freq=="annual",
                          inc_type=="Disposable", tenure=="All", unit=="Household") %>% 
  group_by(area) %>% 
  mutate(source_max=lead(date),
         area=gsub(" Region","",area))

Reg_benchmarks2 <- filter(Reg_benchmarks, inc_type=="Disposable", tenure=="All", unit=="Household") %>% 
  mutate(area=gsub(" Region","",area))

p6 <- ggplot(HUD_mod_inc_reg, aes(date,value))+
  geom_rect(aes(xmin = date, xmax = source_max, ymin = -Inf, ymax = Inf, fill = source), alpha=.4)+
  geom_line(colour=palc[1])+
  geom_point(data=Reg_benchmarks2, aes(y=value, shape="benchmarks"), size=2)+
  facet_wrap(~area)+
  theme_ab()+
  scale_colour_manual(values=palc,name="")+
  scale_fill_manual(values=c("grey85","grey55","#00a0d5"), name="")+
  scale_size_manual(values=c(.7,1,.2))+
  scale_y_continuous(labels=dollar)+
  scale_shape_manual(values=c(19,3), name="")+
  labs(title="Modelled timeseries: Regional Council",
       subtitle="Median annual household disposable income, for all households",
       y="", x="",
       caption="Source: Stats NZ and author's calculations")+
  theme(legend.position="bottom",
        axis.text.x=element_text(angle = 90))+
  guides(fill = guide_legend(ncol=1),
         colour = guide_legend(ncol=1), override.aes = list(alpha = 1))

p6

```

### Figure 7 {-}

```{r, fig.width = 10, fig.height = 14}

#7. Plot of territorial authority timeseries

HUD_mod_inc_ta <- filter(HUD_mod_inc, level=="Territorial Local Authority", inc_freq=="annual",
                          inc_type=="Disposable", tenure=="All", unit=="Household",
                         area !="Chatham Islands Territory") %>% 
  group_by(area) %>% 
  mutate(source_max=lead(date))

TA_benchmarks2 <- filter(TA_benchmarks, inc_type=="Disposable", tenure=="All", unit=="Household") 

p7 <- ggplot(HUD_mod_inc_ta, aes(date,value))+
  geom_rect(aes(xmin = date, xmax = source_max, ymin = -Inf, ymax = Inf, fill = source), alpha=.41)+
  geom_line(colour=palc[1])+
  geom_point(data=TA_benchmarks2, aes(y=value, shape="benchmarks"), size=1)+
  facet_wrap(~area, labeller = labeller(area = label_wrap_gen(18)), ncol=5)+
  theme_ab()+
  scale_colour_manual(values=palc,name="")+
  scale_fill_manual(values=c("grey85","grey55","#00a0d5"), name="")+
  scale_size_manual(values=c(.7,1,.2))+
  scale_y_continuous(labels=dollar)+
  scale_shape_manual(values=19, name="")+
  labs(title="Modelled timeseries: Territorial Local Authority",
       subtitle="Median annual household disposable income, for all households",
       y="", x="",
       caption="Source: Stats NZ and author's calculations")+
  theme(legend.position="bottom",
        axis.text.x=element_text(angle = 90))+
  guides(fill = guide_legend(ncol=1),
         colour = guide_legend(ncol=1), override.aes = list(alpha = 1))

p7

```


```{r, fig.width = 10, fig.height = 14}
#8. Create monthly series

#8.1. Create "missing months' from quarterly series

months <- seq(min(HUD_mod_inc$date)+days(1),max(HUD_mod_inc$date)+days(1), by="month")

months_missing <- as.Date(setdiff(months,unique(HUD_mod_inc$date)+days(1)))-days(1)

months_missing_df <- data.frame(date=months_missing)

#8.2 Add missing dates, then linear interpolation of values

HUD_mod_inc_monthly <- HUD_mod_inc %>%
  ungroup() %>% 
  filter(area != "Chatham Islands Territory") %>% 
  select(-c(date,value,source)) %>% 
  unique() %>% 
  full_join(months_missing_df, by = character()) %>% #cross-join to create all possible combinations of income types
  mutate(value=NA,
         source=NA) %>% 
  rbind(filter(HUD_mod_inc, area != "Chatham Islands Territory")) %>% 
  arrange(date) %>%
  group_by(area,unit,inc_type,tenure,inc_freq) %>% 
  fill(source, .direction="up") %>%
  mutate(value=na.approx(value))

#8.3 Output the final series
write.csv(HUD_mod_inc_monthly,"Modelled_income.csv",row.names=F)

#8.4 check monthly series compared with quarterly

HUD_mod_inc_ta2 <- HUD_mod_inc_ta %>% 
  mutate(freq="quarterly") %>% 
  filter(date >= as.Date("2010-01-01"))

TA_benchmarks3 <- TA_benchmarks2 %>% 
  filter(date >= as.Date("2010-01-01"))

HUD_mod_inc_ta_monthly <- filter(HUD_mod_inc_monthly, level=="Territorial Local Authority", inc_freq=="annual",
                          inc_type=="Disposable", tenure=="All", unit=="Household",
                         area !="Chatham Islands Territory") %>% 
  group_by(area) %>% 
  mutate(source_max=lead(date),
         freq="monthly") %>% 
  filter(date >= as.Date("2010-01-01"))

p8 <- ggplot(HUD_mod_inc_ta2, aes(date,value))+
  geom_rect(aes(xmin = date, xmax = source_max, ymin = -Inf, ymax = Inf, fill = source), alpha=.4)+
  geom_line(aes(colour=freq))+
  geom_line(data=HUD_mod_inc_ta_monthly, aes(colour=freq))+
  geom_point(data=TA_benchmarks3, aes(y=value, shape="benchmarks"), size=1)+
  facet_wrap(~area, labeller = labeller(area = label_wrap_gen(18)), ncol=5)+
  theme_ab()+
  scale_colour_manual(values=palc,name="")+
  scale_fill_manual(values=c("grey85","grey70","grey55"), name="")+
  scale_size_manual(values=c(.7,1,.2))+
  scale_y_continuous(labels=dollar)+
  scale_shape_manual(values=19, name="")+
  labs(title="Modelled timeseries: Territorial Local Authority",
       subtitle="Median annual household disposable income, for all households",
       y="", x="",
       caption="Source: Stats NZ and author's calculations")+
  theme(legend.position="bottom",
        axis.text.x=element_text(angle = 90))+
  guides(fill = guide_legend(ncol=1),
         colour = guide_legend(ncol=1), override.aes = list(alpha = 1))

# p8

```

# Appendix {-}

## Comparisons with other data sources {-}

### Figure A1: Comparison of modelled timeseries with HES and HLFS {-}

```{r}

#9. Compare modelled timeseries with Household Economic Survey (HES) estimates

HUD_mod_inc_NZ_hhdis <- filter(HUD_mod_inc_a, area=="New Zealand", tenure=="All", unit=="Household") %>%
  group_by(tenure,unit) %>% 
  mutate(source_max=lead(date))


NZ_benchmarks_hhdis <- NZ_benchmarks %>% 
  filter(tenure=="All", unit=="Household")

#9.1 Household Economic Survey (HES) estimates 2007-2018

HES_inc2 <- HES_hh %>% 
  rbind(HES_per) %>% 
  gather(inc_type,value,-c(Year,RegCouncil,HouseholdTenure,unit)) %>% 
  filter(RegCouncil == "Total") %>% 
  group_by(inc_type,HouseholdTenure,unit) %>% 
  mutate(date=as.Date(paste0(Year,"-03-31")),
    x=1:n(),
         value_smooth=predict(loess(value ~ x, span=0.5)),
    tenure=recode(HouseholdTenure,
                  "Dwelling not owned, rent payments made"="Renters",
                  "Total"="All")) %>% #smooth series
  filter(inc_type %in% c("MedianTotalRegular","MedianDispIncome")) %>% 
  ungroup() %>% 
  select(unit,tenure,inc_type,value,date,value_smooth) %>% 
  gather(trend_type,value,-c(unit,tenure,inc_type,date)) %>% 
  mutate(trend_type=recode(trend_type,
                                  'value'="actual",
                                  'value_smooth'="smoothed"),
         inc_type=recode(inc_type,
                         "MedianTotalRegular"="Gross",
                         "MedianDispIncome"="Disposable"))

HES_inc_hhdis <- HES_inc2 %>% 
  filter(tenure=="All", unit=="Household",
         trend_type=="actual") %>% 
  mutate(date=date-90) %>% 
  select(date,inc_type,value)

#9.2 Add more recent HES estimates

#Stats NZ data 2020

download.file("https://www.stats.govt.nz/assets/Uploads/Household-income-and-housing-cost-statistics/Household-income-and-housing-cost-statistics-Year-ended-June-2020/Download-data-corrected/Household-income-and-housing-cost-statistics-year-ended-june-2020-corrected-CSV.zip",
              destfile = "housing_costs_2020.zip")

HES_inc_2020 <- read.csv(unz("housing_costs_2020.zip","inc_t4_datafile_2020.csv")) %>% 
  filter( MsCode %in% c("M002"), #median
          IncCode %in% c("I001","I002"), #gross or disposable
          RegCode %in% c("R013")) %>% #NZ
  select(Estimate,IncCode) %>% 
  mutate(date=as.Date("2020-01-01"),
         inc_type=ifelse(IncCode=="I001","Gross","Disposable"),
         value=Estimate) %>% 
  select(date,inc_type,value)


#Stats NZ data 2021

download.file("https://www.stats.govt.nz/assets/Uploads/Household-income-and-housing-cost-statistics/Household-income-and-housing-cost-statistics-Year-ended-June-2021/Download-data/household-income-and-housing-cost-statistics-year-ended-june-2021-CSV.zip",
              destfile = "housing_costs_2021.zip")

HES_inc_2021 <- read.csv(unz("housing_costs_2021.zip","inc_t4_datafile_2021.csv")) %>% 
  filter( MsCode %in% c("M002"), #median
          IncCode %in% c("I001","I002"), #gross or disposable
          RegCode %in% c("R013")) %>% #NZ
  select(Estimate,IncCode) %>% 
  mutate(date=as.Date("2021-01-01"),
         inc_type=ifelse(IncCode=="I001","Gross","Disposable"),
         value=Estimate) %>% 
  select(date,inc_type,value)

#Revised timeseries (only available for NZ total, not regional councils)

HES_inc_REV <- read.csv(unz("housing_costs_2021.zip","inc_t3_datafile_2021.csv")) %>% 
  filter( MsCode %in% c("M002"), #median
          IncCode %in% c("I001A","I002A")) %>% #gross or disposable
  select(Estimate,IncCode, Year) %>% 
  mutate(date=as.Date(paste0(Year,"-01-01")),
         inc_type=ifelse(IncCode=="I001A","Gross","Disposable"),
         value=Estimate,
         tenure="All",
         unit="Household") %>% 
  select(date,inc_type,value, tenure, unit) %>% 
  filter(date !=as.Date("2021-01-01")) #exclude most recent estimate as not a revision

HES_inc_hhdis2 <- HES_inc_hhdis %>% 
  rbind(HES_inc_2020) %>%  #add 2020 data
  rbind(HES_inc_2021) #add 2021 data


#Add revised HES estimates

download.file("https://www.stats.govt.nz/assets/Uploads/household-income-and-housing-cost-statistics-year-ended-june-2019-csv.zip",
              destfile = "housing_costs_2019.zip")

HES_inc_2019_rev <- read.csv(unz("housing_costs_2019.zip","inc-t2-datafile-csv.csv")) %>% 
  filter( MsCode %in% c("M002"), #median
          IncCode %in% c("I001A", "I002A")) %>% 
  mutate(date=as.Date(paste0(Year,"-01-01")),
         vintage="2020",
         inc_type=ifelse(IncCode=="I001A","Gross","Disposable")) %>% 
  select(date,vintage,Estimate, inc_type)

HES_inc_2020_rev <- read.csv(unz("housing_costs_2020.zip","inc_t3_datafile_2020.csv")) %>% 
  filter( MsCode %in% c("M002"), #median
          IncCode %in% c("I001A", "I002A")) %>% 
  mutate(date=as.Date(paste0(Year,"-01-01")),
         vintage="2021",
         inc_type=ifelse(IncCode=="I001A","Gross","Disposable")) %>% 
  select(date,vintage,Estimate, inc_type)

HES_inc_2021_rev <- read.csv(unz("housing_costs_2021.zip","inc_t3_datafile_2021.csv")) %>% 
  filter( MsCode %in% c("M002"), #median
          IncCode %in% c("I001A", "I002A")) %>% 
  mutate(date=as.Date(paste0(Year,"-01-01")),
         vintage="2022",
         inc_type=ifelse(IncCode=="I001A","Gross","Disposable")) %>% 
  select(date,vintage,Estimate, inc_type)

HES_inc_rev <- rbind(HES_inc_2019_rev, HES_inc_2020_rev, HES_inc_2021_rev)


#Add revised HES to chart

HES_inc_rev2 <- rbind(HES_inc_2020_rev, HES_inc_2021_rev) %>%
  filter(date == as.Date("2019-01-01") & vintage==2021 |
         date == as.Date("2020-01-01") & vintage==2022) %>% 
  rename(value=Estimate)

#Add HLFS for comparison

HLFS_NZ <- read.csv("../data/HLFS_NZ.csv", stringsAsFactors = F) %>% 
  mutate(date=as.Date(paste0(year,"-06-30")),
         value=med_inc*52)

#9.3 Plot the data

pA1 <- ggplot(HUD_mod_inc_NZ_hhdis, aes(date,value))+
  geom_point(data=NZ_benchmarks_hhdis, aes(shape="benchmarks",colour=inc_type), size=2)+
  geom_line(aes(colour=inc_type))+
  geom_line(data=HLFS_NZ, aes(colour="Gross"), size=.5, alpha=.5)+
  geom_point(data=HLFS_NZ, aes(colour="Gross",shape="HLFS"), size=1, alpha=.8)+
  geom_line(data=HES_inc_hhdis2, aes(linetype="Household Economic Survey", colour=inc_type), show.legend = F)+
  geom_point(data=HES_inc_hhdis2, aes(shape="Household Economic Survey, first published", colour=inc_type))+
  geom_point(data=HES_inc_rev2, aes(shape="Household Economic Survey, revised", colour=inc_type))+
  geom_point(data=HES_inc_REV, aes(shape="Household Economic Survey, revised", colour=inc_type))+
  geom_line(data=HES_inc_REV, aes(linetype="Household Economic Survey, revised", colour=inc_type), show.legend = F)+
  geom_vline(xintercept=as.Date("2018-01-01"), colour=palc[4])+
  annotate("text", x = as.Date("2018-01-01"), y = 50000, label = " new \n HES-admin \n methodology", colour=palc[4], hjust="left")+
  # facet_grid(tenure~unit)+
  theme_ab()+
  scale_colour_manual(values=palc,name="")+
  scale_fill_manual(values=c("grey85","grey70","grey55"), name="")+
  scale_size_manual(values=c(.7,1,.2))+
  scale_linetype_manual(values=c("dotted","dashed"))+
  scale_y_continuous(labels=dollar)+
  scale_shape_manual(values=c(19,0,3,18), name="")+
  labs(title="Modelled timeseries compared with Household Economic Survey",
       subtitle="New Zealand, Median household income",
       y="", x="",
       caption="Source: Stats NZ and author's calculations")+
  theme(legend.position="bottom")+
  guides(fill = guide_legend(ncol=1),
         shape = guide_legend(ncol=1),
         colour = guide_legend(ncol=1), override.aes = list(alpha = 1))

pA1

```

Figure A1 shows a comparison of the final modelled income timeseries for New Zealand with the model benchmarks and estimates from the Household Economic Survey.[^2] 

[^2]: Household gross (before tax) 'median income from all sources collected' from the Household Labour Force Survey (HLFS), annualised from a weekly amount, is shown as a further comparison.

A methodology change in the income estimator used in the Household Economic Survey for estimates from the year end June 2019 complicates timeseries comparisons. The new methodology uses income data from linked government administrative records to replace some sources of income for eligible individuals (labelled as "HES-admin" in Figure A1). According to Stats NZ's [*Changes to the household economic survey 2018/19*](https://www.stats.govt.nz/methods/changes-to-the-household-economic-survey-201819)[^3] the new estimates of household income will be more precise, so the revised estimates are used in the calibration factors shown in Table 2. Figure A1 also shows more recent revisions to the first published estimates for the year to June 2019 and 2020 (these are due to the availability of additional administrative data, and revisions to population estimates used to calculate the survey weights).

[^3]: https://www.stats.govt.nz/methods/changes-to-the-household-economic-survey-201819

Figure A2 shows a comparison of the final modelled income timeseries for selected regional councils with estimates from the Household Economic Survey. Unfortunately, backcast series using the new methodology are not available by region, so there is a break in the timeseries between the HES estimates between the year to June 2018 and year to June 2019

### Figure A2: Regional comparison of modelled timeseries with HES {-}

```{r, fig.height=8}

#10. Compare modelled timeseries with Household Economic Survey (HES) estimates, by Regional Council

#10.1 HES series 2007-2019
HES_hh_reg_a <- HES_hh %>% 
  filter(HouseholdTenure=="Total") %>% 
  mutate(date=as.Date(paste0(Year,"-03-31"))) %>% 
  select(RegCouncil,date,MedianDispIncome) %>% 
  rename(value=MedianDispIncome,
         area=RegCouncil) %>% 
  filter(area %in% unique(HUD_mod_inc_reg$area)) %>% 
  select(date,value,area) %>% 
  mutate(inc_type="Disposable")

HES_hh_reg_b <- HES_hh %>% 
  filter(HouseholdTenure=="Total") %>% 
  mutate(date=as.Date(paste0(Year,"-03-31"))) %>% 
  select(RegCouncil,date,MedianTotalRegular) %>% 
  rename(value=MedianTotalRegular,
         area=RegCouncil) %>% 
  filter(area %in% unique(HUD_mod_inc_reg$area)) %>% 
  select(date,value,area) %>% 
  mutate(inc_type="Gross")

HES_hh_reg <- rbind(HES_hh_reg_a, HES_hh_reg_b)

#10.2 HES2020 series
HES_inc_2020_r <- read.csv(unz("housing_costs_2020.zip","inc_t4_datafile_2020.csv")) %>% 
  filter( MsCode %in% c("M002"), #median
          IncCode %in% c("I001","I002"), #gross or disposable
          RegCode %in% c("R001","R002","R003",
                         "R004","R006","R007",
                         "R008","R010","R011",
                         "R012")) %>% #comparable regions
  select(Estimate,RegCode, IncCode) %>% 
  mutate(date=as.Date("2020-01-01"),
         inc_type=ifelse(IncCode=="I001","Gross","Disposable"),
         value=Estimate,
         area=recode(RegCode,
                     "R001"="Northland",
                     "R002"="Auckland",
                     "R003"="Waikato",
                     "R004"="Bay of Plenty",
                     "R006"="Taranaki",
                     "R007"="Manawatu-Whanganui",
                     "R008"="Wellington",
                     "R010"="Canterbury",
                     "R011"="Otago",
                     "R012"="Southland" )) %>% 
  select(date,value,area,inc_type)

#HES2021 series

HES_inc_2021_r <- read.csv(unz("housing_costs_2021.zip","inc_t4_datafile_2021.csv")) %>% 
  filter( MsCode %in% c("M002"), #median
          IncCode %in% c("I001","I002"), #gross or disposable
          RegCode %in% c("R001","R002","R003",
                         "R004","R006","R007",
                         "R008","R010","R011",
                         "R012")) %>% #comparable regions
  select(Estimate,RegCode, IncCode) %>% 
  mutate(date=as.Date("2021-01-01"),
         inc_type=ifelse(IncCode=="I001","Gross","Disposable"),
         value=Estimate,
         area=recode(RegCode,
                     "R001"="Northland",
                     "R002"="Auckland",
                     "R003"="Waikato",
                     "R004"="Bay of Plenty",
                     "R006"="Taranaki",
                     "R007"="Manawatu-Whanganui",
                     "R008"="Wellington",
                     "R010"="Canterbury",
                     "R011"="Otago",
                     "R012"="Southland" )) %>% 
  select(date,value,area,inc_type)


HES_hh_reg2 <- HES_hh_reg %>% 
  rbind(HES_inc_2020_r) %>% 
  rbind(HES_inc_2021_r)


HUD_mod_inc_reg2 <- filter(HUD_mod_inc, level=="Regional Council", inc_freq=="annual",
                          # inc_type=="Disposable", 
                          tenure=="All", unit=="Household") %>% 
  group_by(area) %>% 
  mutate(source_max=lead(date),
         area=gsub(" Region","",area)) %>% 
  filter(area %in% unique(HES_hh_reg$area)) 

# HUD_mod_inc_reg2 <- HUD_mod_inc_reg %>% 
#   filter(area %in% unique(HES_hh_reg$area)) 

Reg_benchmarks3 <- filter(Reg_benchmarks, 
                          # inc_type=="Disposable", 
                          tenure=="All", unit=="Household") %>% 
  mutate(area=gsub(" Region","",area)) %>% 
  filter(area %in% unique(HES_hh_reg$area))

# Reg_benchmarks3 <- Reg_benchmarks2 %>% 
#   filter(area %in% unique(HES_hh_reg$area))

#HLFS data for comparison

HLFS <- read.csv("../data/HLFS_med_income.csv", na.strings=c("","NA"), stringsAsFactors = F) %>% 
  fill(Year) %>% 
  filter(!(Region %in% c("Gisborne/Hawkes Bay Regions","Tasman/Nelson/Marlborough/West Coast Regions")),
         !is.na(Income)) %>% 
  mutate(date=as.Date(paste0(Year,"-06-30")),
         area=gsub(" Region","",Region),
         area=recode(area,
                     "Manawatu-Wanganui"="Manawatu-Whanganui"),
         value=Income*52) 

pA2 <- ggplot(HUD_mod_inc_reg2, aes(date,value, colour=inc_type))+
  geom_line(aes(linetype="model"))+
  geom_point(data=Reg_benchmarks3, aes(y=value, shape="benchmarks"), size=2)+
  geom_point(data=HES_hh_reg2, aes(y=value, shape="Household Economic Survey"), size=2)+
  geom_point(data=HLFS, aes(y=value, shape="HLFS (Gross)", colour="Gross", linetype="survey"), size=1)+
  geom_line(data=HES_hh_reg2, aes(y=value, linetype="survey"), size=.3)+
  geom_vline(xintercept=as.Date("2018-08-01"), colour=palc[4])+
  annotate("text", x = as.Date("2018-08-01"), y = 40000, label = " new HES \n methods", colour=palc[4], hjust="left", size=2.5)+
  facet_wrap(~area, ncol=2)+
  theme_ab()+
  scale_colour_manual(values=palc,name="")+
  scale_fill_manual(values=c("grey85","grey70","grey55"), name="")+
  scale_size_manual(values=c(.7,1,.2))+
  scale_linetype_manual(values=c("solid","dotted"), name="")+
  scale_y_continuous(labels=dollar)+
  scale_shape_manual(values=c(19,0,3), name="")+
  labs(title="Modelled timeseries: Regional Council",
       subtitle="Median annual household disposable income, for all households",
       y="", x="",
       caption="Source: Stats NZ and author's calculations")+
  theme(legend.position="bottom")+
  guides(fill = guide_legend(ncol=1),
         shape = guide_legend(ncol=1),
         linetype = guide_legend(ncol=1),
         colour = guide_legend(ncol=1), override.aes = list(alpha = 1))

pA2


```

### Figure A3: Comparison of wage measures {-}

```{r}

download.file("https://www.stats.govt.nz/assets/Uploads/Labour-market-statistics/Labour-market-statistics-March-2022-quarter/Download-data/labour-market-statistics-march-2022-quarter-csv.zip",
              destfile = "lab.zip")

lci <- read.csv(unz("lab.zip", "labour-market-statistics-march-2022-quarter-csv/lci-mar22qtr-csv.csv"), stringsAsFactors = F) 

qes <- read.csv(unz("lab.zip", "labour-market-statistics-march-2022-quarter-csv/qes-mar22qtr-csv.csv"), stringsAsFactors = F)


lci_all <- lci %>% 
  filter(Group == "All Sectors Combined, All Salary and Wage Rates (Base: June 2017 qtr (=1000))") %>% 
  mutate(date=as.Date(paste0(Period,".01"),"%Y.%m.%d"),
         date=date_to_qtr(date)) %>% 
  arrange(date) %>% 
  mutate(ann.change=Data_value/lag(Data_value,4)-1,
         series="LCI") %>% 
  filter(!is.na(ann.change)) %>% 
  select(ann.change, Data_value, date, series)

qes_con2 <- qes %>% 
  filter(Group == "Average Hourly Earnings by Industry (ANZSIC06) and Sex",
         Series_title_1 == "Total All Industries",
         Series_title_2== "Total Both Sexes",
         Series_title_3 == "Ordinary Time Hourly" 
  ) %>% 
  mutate(date=as.Date(paste0(Period,".01"),"%Y.%m.%d"),
         date=date_to_qtr(date)) %>% 
  group_by(Series_title_2) %>% 
  arrange(date) %>% 
  mutate(ann.change=Data_value/lag(Data_value,4)-1,
         series="QES") %>% 
  ungroup() %>% 
  select(ann.change, Data_value, date, series)

#average earnings from business employment survey

bus_emp <- read.csv(unz("bus_emp_mar22.zip", "machine-readable-business-employment-data-mar-2022-quarter.csv"), stringsAsFactors = F)

bus_emp_con_ave_earn <- bus_emp %>% 
  filter(Group == "Industry by employment variable",
         Series_title_2 == "Total Industry",
         Series_title_3 == "Actual") %>% 
  mutate(date=as.Date(paste0(Period,".01"),"%Y.%m.%d"),
         date=date_to_qtr(date)) %>% 
  select(date, Data_value, Series_title_1) %>% 
  group_by(date) %>% 
  arrange(date) %>% 
  mutate(Data_value=Data_value*1000000/lag(Data_value,1),
         Series_title_1="Earnings per job") %>% 
  filter(!is.na(Data_value)) %>% 
  ungroup() %>% 
  arrange(date) %>% 
  mutate(Data_value=Data_value+lag(Data_value,1)+lag(Data_value,2)+
           lag(Data_value,3),
    ann.change=Data_value/lag(Data_value,4)-1) %>% 
  filter(!is.na(ann.change)) %>% 
  mutate(series="BEI")%>% 
  select(ann.change, Data_value, date, series)


#LEED (annualised)

leed_ta <- read.csv("../data/LEED_table37.csv", na.strings=c("","NA"), stringsAsFactors = F)[,c(1:2,7)] %>% 
  fill(Quarter) %>%
  rename(area=Territorial.authority,
         med_inc=Median.earnings.of.full.quarters.jobs) %>% 
  filter(med_inc!="..") %>% 
  mutate(area=str_trim(area),
         area=gsub("district","District",area),
         area=gsub("city","City",area),
         area=gsub(" \\(modified\\)","",area),
         area=gsub("Total territorial authority","New Zealand",area),
         area=gsub("Wanganui District","Whanganui District",area),
         med_inc=as.numeric(med_inc),
         date=as.Date(paste0(Quarter,"-01"),format="%b-%y-%d"),
         date=date_to_qtr(date),
         source="LEED",
         level=ifelse(area=="New Zealand","New Zealand","Territorial Local Authority")) %>% 
  select(-Quarter) %>% 
  group_by(area) %>% 
  arrange(date) %>% 
  mutate(med_inc=med_inc+lag(med_inc,1)+lag(med_inc,2)+lag(med_inc,3)) %>% #annual income
  ungroup()

leed_nz <- leed_ta %>% 
  filter(area == "New Zealand") %>% 
  select(date,med_inc) %>% 
  rename(Data_value=med_inc) %>% 
  arrange(date) %>% 
  mutate(ann.change=Data_value/lag(Data_value,4)-1) %>% 
  filter(!is.na(ann.change)) %>% 
  mutate(series="LEED")%>% 
  select(ann.change, Data_value, date, series)


NZ_benchmarks_hhdis_c <- HUD_mod_inc_monthly %>% 
  filter(tenure == "All",
         unit == "Household",
         inc_freq == "annual",
    inc_type == "Disposable",
    area == "New Zealand") %>% 
  ungroup() %>% 
  arrange(date) %>% 
  rename(Data_value=value) %>% 
  mutate(ann.change=Data_value/lag(Data_value,12)-1,
         series="Modelled") %>% 
  select(ann.change, Data_value, date, series)
  

wagecombo <- rbind(qes_con2, lci_all,bus_emp_con_ave_earn, leed_nz, NZ_benchmarks_hhdis_c) %>% 
  filter(date >= as.Date("2000-03-31"))

pA3 <- ggplot(wagecombo, aes(date, ann.change, colour=series))+
  geom_line()+
  theme_ab()+
  scale_y_continuous(labels=percent_format(accuracy=0.1))+
  scale_colour_manual(values=palc2, name="")+
  labs(title="Salary and wage increases \ncompared with modelled household income",
       subtitle="Annual change",
       x="",y="",
       caption="Stats NZ, \nLabour Cost Index (LCI), Quarterly Employment Survey (QES), \nBusiness Employment Indicators (BEI), Linked Employer-Employee Data (LEED)")+
  geom_text(data=filter(wagecombo, !(series %in% c("Modelled","BEI")) & date==max(date)), aes(label=paste0("  ", round(ann.change*100,1), "% ", series), hjust="left"), show.legend = F, size=4)+
  geom_text(data=filter(wagecombo, series=="Modelled" & date==max(date)), aes(label=paste0("  ", round(ann.change*100,1), "% ", series), hjust="left", vjust=1), show.legend = F, size=4)+
  geom_text(data=filter(wagecombo, series=="BEI" & date==max(date)), aes(label=paste0("  ", round(ann.change*100,1), "% ", series), hjust="left", vjust=-.5), show.legend = F, size=4)+
    scale_x_date(breaks=c(min(wagecombo$date),as.Date("2010-01-01"), as.Date("2020-01-01"),max(wagecombo$date)),
                 minor_breaks=c(min(wagecombo$date),as.Date("2010-01-01"), as.Date("2020-01-01"),max(wagecombo$date)),
                limits=c(min(wagecombo$date),max(wagecombo$date)+1500),
               labels=c(paste0(year(min(wagecombo$date))," \n",lubridate::month(min(wagecombo$date),label=T)),
                        paste0(year(as.Date("2010-01-01"))," \n",lubridate::month(as.Date("2010-01-01"),label=T)),
                        paste0(year(as.Date("2020-01-01"))," \n",lubridate::month(as.Date("2020-01-01"),label=T)),
                        paste0(year(max(wagecombo$date))," \n",lubridate::month(max(wagecombo$date),label=T))))+
  theme(legend.position = "bottom")

pA3

```

\newpage

## Links to input data sources {-}

### Household Economic Survey (HES) {-}

**Years ending June 2007--2019**

https://catalogue.data.govt.nz/dataset/hes-gross-and-disposable-income-timeseries-2006-07-2018-19

**Years ending June 2007--2019** (revised data for 2007-18)

https://www.stats.govt.nz/information-releases/household-income-and-housing-cost-statistics-year-ended-june-2019

**Years ending June 2019--2020** (revised data for 2019) 
https://www.stats.govt.nz/information-releases/household-income-and-housing-cost-statistics-year-ended-june-2020

**Years ending June 2019--2021** (revised data for 2020)
https://www.stats.govt.nz/information-releases/household-income-and-housing-cost-statistics-year-ended-june-2021

### Censuses of Population and Dwellings (Census) {-}

https://catalogue.data.govt.nz/dataset/census-household-and-personal-income-series

### Linked Employer-Employee Data (LEED) {-}

https://www.stats.govt.nz/information-releases/linked-employer-employee-data-march-2021-quarter-nz-stat-tables

**By Regional Council**

https://nzdotstat.stats.govt.nz/wbos/Index.aspx?DataSetCode=TABLECODE7005

**by Territorial Authority**

https://nzdotstat.stats.govt.nz/wbos/Index.aspx?DataSetCode=TABLECODE7037

### Business Employment Data {-}

https://www.stats.govt.nz/information-releases/business-employment-data-march-2022-quarter/


## Links to comparison data sources {-}

### Household Labour Force Data {-}

https://nzdotstat.stats.govt.nz/wbos/Index.aspx?DataSetCode=TABLECODE7480

### Labour Market Statistics {-}

Labour Cost Index (LCI) and Quarterly Employment Survey (QES)

https://www.stats.govt.nz/information-releases/labour-market-statistics-march-2022-quarter/
